Staged Abstract Interpreters

### 2019-04-06 paper/10_intro.tex
some number and typos

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

Now to concretely illustrate the idea of Futamura projection, consider that we
have an operational interpreter @eval@ of some language:

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe up to 17x times faster
than the unstage abstract interpreter. However, we do not lose any advantages of
this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of
static analysis after staging.

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be given explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

Now to concretely illustrate the idea of Futamura projection, consider that we
have an interpreter @eval@ of some language:

@@

Now that we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

We show that these well-known abstractions can be combined smoothly, and more
importantly for free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe that the staged version is 
11x times as fast as the unstage abstract interpreter. 
However, we do not lose any advantages of this high-level programming style 
(e.g., modularity and extensibility); the meaning of the program also remains 
the same, particularly, the soundness of static analysis after staging.

### 2019-04-06 paper/70_eval.tex
evaluation
some number and typos

To evaluate how the performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows that the staged versions are averagely 6 times as fast of the unstaged
version, as strong evidence of our \textit{abstraction without regret}
approach.

\paragraph{Implementation and Evaluation Environment}
We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis focus on lambda terms in functional languages. A front-end
desugars Scheme to a small core language that makes the analyzer easy to
implement. As described in Section \ref{cfa}, we implement a monovariant
0-CFA-like analysis, \todo{a context-sensitive $1$-CFA-like analysis, both their
  versions with and without store-widening}. We also implement several
optimizations mentioned in Section \ref{staged_ds} (less than ~30 LOC in total).

@@

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening} &  \multicolumn{3}{c}{w/ store-widening}  \\
    \midrule
    @fib@               & 32   & 0.013      & 0.002      & 650.00\%  & & &        \\
    @rsa@               & 451  & 0.232      & 0.042      & 552.38\%  & & &        \\
    @church@            & 120  & 67.342     & 18.917     & 355.99\%  & & &        \\
    @fermat@            & 310  & 0.024      & 0.004      & 600.00\%  & & &        \\
    @mbrotZ@            & 331  & 0.763      & 0.127      & 600.79\%  & & &        \\
    @lattice@           & 609  & 31.874     & 8.816      & 361.59\%  & & &       \\
    @kcfa-worst-16@     & 182  & 0.045      & 0.006      & 750.00\%  & & &      \\
    @kcfa-worst-32@     & 358  & 0.284      & 0.022      & 1290.91\% & & &      \\
    @kcfa-worst-64@     & 774  & 2.18       & 0.129      & 1689.92\% & & &      \\
    @solovay-strassen@  & 523  & 5.094      & 1.189      & 428.43\%  & & &       \\
    @regex@             & 550  & -          & -          & -         & & &       \\
    @matrix@            & 1891 & -          & -          & -         & & &       \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on 0CFA without store-widening. The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis (in seconds). The last column $\frac{\text{unstaged}}{\text{staged}}$
shows the improvement from staging. The dash - in some cells indicates timeout.
\todo{0cfa, 1cfa, store-widen}
As we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks, especially for the @kcfa-worst-@$n$ series.
We observe the most improvement of 17x times faster than the unstaged version on
@kcfa-worst-64@. On average, the staging produces code that runs 7x times faster
than the unstaged analyzer.

>>>>

To evaluate how the performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows that the staged versions are averagely 11x times as fast as the unstaged
version, as strong evidence of our \textit{abstraction without regret}
approach.

\paragraph{Implementation and Evaluation Environment}
We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis focus on lambda terms in functional languages. A front-end
desugars Scheme to a small core language that makes the analyzer easy to
implement. As described in Section \ref{cfa}, we implement two monovariant
0-CFA-like analysis, one is equipped with store-widening, the other does not.
We also implement several optimizations mentioned in Section \ref{staged_ds} 
(less than ~30 LOC in total).

@@

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening (sec.)}  &  \multicolumn{3}{c}{w/ store-widening (ms)}\\
    \midrule
    @fib@               & 32   & 0.013      & 0.002      & 6.50x      & 1.537 & 0.067  &  22.87x       \\
    @rsa@               & 451  & 0.232      & 0.042      & 5.52x      & 15.113& 1.449 &  10.43x       \\
    @church@            & 120  & 67.342     & 18.917     & 3.56x      & 2.962 & 0.291 &  10.19x       \\
    @fermat@            & 310  & 0.024      & 0.004      & 6.00x      & 7.624 & 0.622 &  12.26x       \\
    @mbrotZ@            & 331  & 0.763      & 0.127      & 6.01x      & 12.229& 0.729 &  16.77x       \\
    @lattice@           & 609  & 31.874     & 8.816      & 3.62x      & 17.943& 0.909 &  19.75x      \\
    @kcfa-worst-16@     & 182  & 0.045      & 0.006      & 7.50x      & 4.658 & 0.785 &  5.93x     \\
    @kcfa-worst-32@     & 358  & 0.284      & 0.022      & 12.91x     & 10.005& 1.680 &  5.95x     \\
    @kcfa-worst-64@     & 774  & 2.18       & 0.129      & 16.90x     & 22.999& 3.872 &  5.94x     \\
    @solovay-strassen@  & 523  & 5.094      & 1.189      & 4.28x      & 19.118& 0.654 &  29.25x      \\
    @regex@             & 550  & -          & -          & -          & 7.133 & 0.624 &  11.43x      \\
    @matrix@            & 1891 & -          & -          & -          & 86.176& 4.592 &  18.76x      \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on 0CFA without and with store-widening. The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis. The column $\frac{\text{unstaged}}{\text{staged}}$
shows the improvement from staging. The dash - in some cells indicates timeout.
As we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks, especially for the @kcfa-worst-@$n$ series.
We observe the most improvement of 29x times faster than the unstaged version on
@solovay-strassen@. On average, the staging produces code that runs 11x times faster
than the unstaged analyzer.

### 2019-04-06 paper/abstract.tex
some number and typos

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application of various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 7x with staging for control-flow analysis.

>>>>

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application of various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 11x with staging for control-flow analysis.
